WASP-4 is Accelerating Towards the Earth

### 2020-01-27 paper/ms.tex
update paper

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.
Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.
A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.
% \citeauthor{baluev_homogeneously_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
% \citeauthor{baluev_2019} found that when they used all the available
% TTV data, the need for a quadratic ephemeris was present.  However,
% they pointed out that if they used lower-precision subsets of the
% available timing data, the necessity for the quadratic term decreased.
% This is not particularly surprising.
% \citeauthor{baluev_homogeneously_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_homogeneously_2019} did not find the claim of a
% period decrease convincing.

@@

{\bf Table~N} lists the transit times we used in our analysis.  We
include data from the peer-reviewed literature for which (a) the
analysis was based on observations of a single transit, (b) the
midpoint was fitted as a free parameter, and (c) the time system
specified both the absolute reference (TDB or UTC) and also whether
any barycentric or heliocentric corrections had been performed.

The majority of times are identical to those we collected in
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m NTT and Danish 1.54m telescopes at La Silla,
and the SAAO 1.0m telescope.

% FIXME : verify with Michael Gillon and Phil Evans that this is true.
A number of additional times were also recently reported by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations.  We included their ``high quality'' transit
times with provenance from the TRAPPIST and El Sauce telescopes, for
which we verified with the original observers that appropriate
time-system corrections had been performed (M.~Gillon, P.~Evans,
priv{.} comm{.}).

The four available occultation data points tabulated by
\citetalias{bouma_wasp4b_2019} are sparse and have neglegible
statistical value due to their large uncertainties, and we forgo
their use in this analysis.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed times minus the calculated
  times assuming a constant orbital period .  Darker points correspond
  to more precise data.  The quadratic ephemeris and its $1\sigma$
  uncertainties are shown in blue.  The binned TESS point (red square)
  is the weighted average of 18 TESS transits and is for display
  purposes only.  The models were fitted to all of the individual
  transit times.
  \label{fig:times}
	}
\end{figure}

@@

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model. 
The best-fitting constant-period model has $\chi^2 = XXX$ and YY
degrees of freedom.  The best-fitting
quadratic model has $\chi^2 = YYY$ and ZZ degrees of freedom. 
The difference in the BIC \citep{kass_bayes_1995} between the linear
and quadratic and models is $\Delta {\rm BIC} = ZZ.Z$, strongly
favoring the quadratic case.

The best-fit period derivative in the quadratic model is
\begin{equation}
\dot{P}
  = - (X.XX \pm 0.XX)\times 10^{-10}
  = - YY.Y \pm Z.Z~{\rm ms}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees with the value reported by \citet{southworth_transit_2019}
after their including an additional 22 archival transit measurements
($\dot{P} = - 9.2 \pm 1.1~{\rm ms}\,{\rm yr}^{-1}$).
It is smaller than the rate of period decrease found by
\citetalias{bouma_wasp4b_2019} ($- 12.6 \pm 1.2~{\rm ms}\,{\rm
yr}^{-1}$)., presumably because of missing coverage filled in by
\citeauthor{southworth_transit_2019}'s observations.
The other best-fit transit timing model parameters are reported in
{\bf Table~X}.

@@

%FIXME: is the purple here from my Pdot, or Southworth's?
%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - YY.Y \pm Z.Z~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the slope from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period
derivative expected from radial velocities is
\begin{align}
  \dot{P}_{\rm RV} &= \frac{\dot{v}_{\rm r} P}{c} \\
  \dot{P}_{\rm RV} &= -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P} = - X.X \pm 1.1~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.
% FIXME
All the
available TTV data show a period decrease of $\approx 8-12$
milliseconds per year~\citep{bouma_wasp-4b_2019,southworth_transit_2019,baluev_homogeneously_2019}.

>>>>

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.
Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.
A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.
% \citeauthor{baluev_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
% \citeauthor{baluev_2019} found that when they used all the available
% TTV data, the need for a quadratic ephemeris was present.  However,
% they pointed out that if they used lower-precision subsets of the
% available timing data, the necessity for the quadratic term decreased.
% This is not particularly surprising.
% \citeauthor{baluev_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_2019} did not find the claim of a
% period decrease convincing.

@@

{\bf Table~N} lists the transit times we used in our analysis.  We
include data from the peer-reviewed literature for which (a) the
analysis was based on observations of a single transit, (b) the
midpoint was fitted as a free parameter, and (c) the time system
specified both the leap second correction (TDB or UTC) and also
whether any barycentric or heliocentric corrections had been
performed.

% FIXME : verify with Michael Gillon that this is true. (P. Evans
% done)
The majority of times are identical to those we collected in
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m NTT and Danish 1.54m telescopes at La Silla,
and the SAAO 1.0m telescope.  Additional timimg
measurements have also recently been reported by \citet{baluev_2019},
based on a homogeneous analysis of archival ground-based observations
taken professional and amateur observers.  We included
twelve of their ``high quality'' transit times acquired by TRAPPIST
(six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had or had not been performed.

The four available occultations tabulated by
\citetalias{bouma_wasp4b_2019} have neglegible statistical value due
to their large uncertainties, and we forgo their use in this analysis.

@@

\subsection{Transits}
\label{sec:transit_analysis}

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed times minus the calculated
  times assuming a constant orbital period.  More opaque points correspond
  to more precise data.  The quadratic ephemeris and its $1\sigma$
  statistical uncertainties are shown in blue.  Neither model fully
  explains the transit residuals (see~\ref{sec:transit_analysis}).
  The binned TESS point (yellow star) is the weighted average of 18
  TESS transits and is for display purposes only.  Its uncertainty is
  slightly larger than the marker size.  The models were fitted to all
  of the individual transit times.
  \label{fig:times}
	}
\end{figure}

@@

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model.  The best-fitting constant-period
model has $\chi^2 = 276$, 91 degrees of freedom, and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has $\chi^2 = 183$, 90
degrees of freedom, and $\chi^2_{\rm red}=2.0$.  From the reduced
$\chi^2$ values, we can surmise that neither model entirely describes
the transit data that we have collected---there must be some
additional source of signal or potentially noise. In
\citet{bouma_wasp4b_2019}, we found that the quadratic model for the
(less complete) transit data gave $\chi^2_{\rm red}=1.0$.  The only
difference between our old and new transit datasets are the addition
of transit times from \citet{southworth_transit_2019} and
\citet{baluev_2019}.  The worsened $\chi^2_{\rm red}$ could reflect
underestimated statistical uncertainties in these or in fact any of
the transit measurements. It could also reflect systematic errors in
the time-systems in which the transit measurements were recorded,
though we have taken every caution against the latter possibility.
% For purposes of simplicity, we simply accept that the additional data
% have effectively increased the scatter in the transit O-C diagram.
The difference in the BIC \citep{kass_bayes_1995} between the linear
and quadratic and models is $\Delta {\rm BIC} = 89.1$, strongly
favoring the quadratic case.

The best-fit period derivative in the quadratic model is
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$3$\sigma$ larger than the rate
of period decrease we found earlier \citep[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of
missing coverage filled in by \citeauthor{southworth_transit_2019}'s
and \citeauthor{baluev_2019}'s observations.  The other best-fit
transit timing model parameters are reported in {\bf Table~X}.

@@

%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the slope from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period
derivative expected from radial velocities is
\begin{align}
  \dot{P}_{\rm RV} &= \frac{\dot{v}_{\rm r} P}{c} \\
  \dot{P}_{\rm RV} &= -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.
